Analysis of pleiotropy of eQTLs

### 2023-01-19 presentation_230120_internal_seminar/intern_sem_230120.tex
ENH

%          However, how do we know if these variants are active in a tissue,
%         or whether these variants false positives due to linkage disequilibrium,
%         or how these variants act molecularly
        Questions:
%
        \begin{itemize}
            \item Correlation between variants. Which is the causal variant?
            \item Which cell types important for disease?
            \item In which tissues are these variants active?
        \end{itemize}
%
        \vfill
%
        Questions partially addressed by looking at eQTLs in different tissues

        \blfootnote{}
        \blfootnote{Cano-Gamez et al. 2020. doi:10.3389/fgene.2020.00424}
    \end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Expression quantitative trait loci (eQTLs)}
%eQTLs are variants that modify gene expression in a given tissue.
%        eQTLs give a potential mechanism (the gene target) and a tissue.
        \begin{columns}
            \begin{column}{0.5\textwidth}
                \begin{center}
                    \includegraphics[width=\textwidth]{../presentation_230120_gold2022_paris/fig/doi_10.3389_fgene.2020.00424_fig4a.jpg}
                \end{center}
            \end{column}
            \begin{column}{0.5\textwidth}

            \end{column}
        \end{columns}
%
        \vfill
%
%        To my knowledge, it is unknown where are the most pleiotropice Eqtls in the genome.
%        This is important because this would give hints to putatitve pleiotropi genes (unknown or unknown) that are regulated by these pleiotropic eQTLs?
        Pleiotropy of eQTLs in genome is currently unknown

@@

        \begin{enumerate}
            \item To look for pleiotropic eQTLs in the genome
            \item To characterise their genomic regions, target genes and the cause of their pleiotropy
        \end{enumerate}
    \end{frame}

@@

        \begin{enumerate}
            \item Annotate eQTLs with GWAS traits using colocalization analysis
            \item Annotate GWAS traits in categories
            \item Aggregate trait categories of eQTLs: Pleiotropic eQTLs
            \item Aggregate pleiotropic eQTLs in genome regions
            \item Analyse pleiotropic eQTLs and regions
        \end{enumerate}
    \end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Annotation of GWAS variants with gene and tissue targets using colocalization analysis}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Data and Algorithms}

        \begin{columns}
            \begin{column}{0.5\textwidth}
                \begin{itemize}
                    \item IEA OpenGWAS database\footnote{https://gwas.mrcieu.ac.uk}: 418 GWAS datasets
                    \item EBI eQTLs database\footnote{https://www.ebi.ac.uk/eqtl}: Processed eQTLs from 127 studies in human.
                    \item CRAN - Package coloc\footnote{https://cran.r-project.org/web/packages/coloc/index.html}: Colocalisation analysis of GWAS/eQTL with one causal variant
                \end{itemize}
            \end{column}
            \begin{column}{0.5\textwidth}
                \begin{center}
                    \includegraphics[width=\textwidth]{../presentation_230120_gold2022_paris/fig/pgen.1004383.g001.png}
                \end{center}
            \end{column}
        \end{columns}

%        \let\thefootnote\relax\footnotetext{$^1$https://gwas.mrcieu.ac.uk}
%        \let\thefootnote\relax\footnotetext{$^2$https://www.ebi.ac.uk/eqtl}
%        \let\thefootnote\relax\footnotetext{$^3$https://cran.r-project.org/web/packages/coloc/index.html}
    \end{frame}

@@

        \begin{itemize}
            \item Colocalization between 293 GWAS and 127 eQTL studies
            \item 138e3 colocalized variants with probability $\geq$ 0.75
            \item Percentage of GWAS loci explained by eQTL: 0$\%$ - 77$\%$ (Average 33$\%$)
        \end{itemize}

        \begin{figure}[!]
            \includegraphics[width=\textwidth]{../presentation_230120_gold2022_paris/fig/coloc_web.png}
        \end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Web portail}

        TODO

@@

    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Winner (Most pleiotroic) eQTLs?}

@@

    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Winner (Most pleiotroic) regions?}

@@

@@

    \section{Conclusions, perspective and acknowledgements} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

        \begin{itemize}
            \item A list of pleiotrtA large set of causal GWAS variants annotated with target genes and tissues
            \item A list of pleiotropic variants and regions
        \end{itemize}

        Outcomes interesting for the GOLD population genotypes
%
        \begin{itemize}
            \item A large set of causal GWAS variants annotated with target genes and tissues
            \item A list of pleiotropic variants and regions
        \end{itemize}
%
        \vfill
%
        What are the gene regulatory properties underlying pleiotropy between complex traits?
%
        \begin{itemize}
            \item Variant pleiotropy is correlated in regions
            \item Enrichment of variants upstream (incl. 5'UTR) and downstream of genes (incl. 3'UTR)
            \item Enrichment of transcription factor and CRM binding
            \item Increased number of associated egenes and etissues
        \end{itemize}

@@

        \begin{itemize}
            \item Causality of eQTL genes on GWAS
            \item Study loci in more
        \end{itemize}

>>>>

%          However, how do we know if these variants are active in a tissue,
%         or whether these variants false positives due to linkage disequilibrium,
%         or how these variants act molecularly
        Questions:\footnote{Cano-Gamez et al. 2020. doi:10.3389/fgene.2020.00424}
%
        \begin{itemize}
            \item Correlation between variants. Which is the causal variant?
            \item Which cell types important for disease?
            \item In which tissues are these variants active?
        \end{itemize}
%
        \vfill
%
        Questions partially addressed by looking at eQTLs in different tissues

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Expression quantitative trait loci (eQTLs)}
%eQTLs are variants that modify gene expression in a given tissue.
%        eQTLs give a potential mechanism (the gene target) and a tissue.

        \begin{center}
            \includegraphics[width=\textwidth]{../presentation_230120_gold2022_paris/fig/doi_10.3389_fgene.2020.00424_fig4a.jpg}
        \end{center}
%
%        To my knowledge, it is unknown where are the most pleiotropice Eqtls in the genome.
%        This is important because this would give hints to putatitve pleiotropi genes (unknown or unknown) that are regulated by these pleiotropic eQTLs?
        Pleiotropy of eQTLs in genome is currently unknown

@@

        \begin{enumerate}
            \item To identify pleiotropic eQTLs in the genome
            \item To characterise their genomic regions, target genes and the cause of their pleiotropy
        \end{enumerate}
    \end{frame}

@@

        \begin{enumerate}
            \item Annotate eQTLs with GWAS traits (Colocalization analysis)
            \item Define categories for GWAS traits
            \item Aggregate categories of eQTLs: Pleiotropic eQTLs
            \item Aggregate pleiotropic eQTLs in genome regions
            \item Explore properties pleiotropic eQTLs and regions
        \end{enumerate}
    \end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Colocalization to annotate eQTLs with GWAS traits}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{IEA OpenGWAS database\footnote{https://gwas.mrcieu.ac.uk}}

        \begin{itemize}
            \item 418 GWAS
            \item Physiological and disease traits
            \item UK biobank, GWAS catalog, Other datasets
        \end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{EBI eQTLs database\footnote{https://www.ebi.ac.uk/eqtl}}

        \begin{itemize}
            \item 127 studies in human from different datasets
            \item Tissues, cell types, stimulated conditions, ...
            \item GTEX, DICE, ...
        \end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{CRAN - Package coloc\footnote{https://cran.r-project.org/web/packages/coloc/index.html}}

        \begin{itemize}
            \item Colocalisation analysis of GWAS/eQTL with one causal variant
        \end{itemize}

        \begin{center}
            \includegraphics[width=\textwidth]{../presentation_230120_gold2022_paris/fig/pgen.1004383.g001.png}
        \end{center}

@@

        \begin{itemize}
            \item Colocalization between 293 GWAS and 127 eQTL studies
            \item 25\% of 10 k leading SNPs (SNP.PP.H4$\geq$0.5), colocalized with at least one eQTL
            \item At SNP.PP.H4$\geq$0.5, 5 k eQTLs and 5 k target genes
            \item 138e3 colocalized variants with probability $\geq$ 0.75
            \item Percentage of GWAS loci explained by eQTL: 0$\%$ - 77$\%$ (Average 33$\%$)
        \end{itemize}

        \begin{figure}[!]
            \includegraphics[width=0.5\textwidth]{fig/histogram_explained.png}\label{fig:perc_explained}
        \end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Most pleiotroic eQTLs}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Most pleiotroic regions}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{The IRF1/IL4 locus}

        \begin{figure}[!]
            \includegraphics[width=0.95\textwidth]{fig/il4}
        \end{figure}

        \begin{itemize}
            \item Sevaral immune genes
            \item GWAS categories: allergic disease; autoimmune disease; body height; breast cancer; cancer; cardiovascular disease; hypertensive disorder; respiratory system disease; skin disease
            \item eQTL biosamples: adipose tissue;aorta;blood;brain;breast; ...
        \end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{The CDKN2A/P16 locus}

        \begin{figure}[!]
            \includegraphics[width=0.95\textwidth]{fig/cdk2na}
        \end{figure}

        \begin{itemize}
            \item Cell cycle regulator
            \item GWAS categories: cancer;cardiovascular disease;hypertensive disorder;joint disease;respiratory system disease;skin cancer
            \item eQTL biosamples: adipose tissue;blood;brain;colon;immune system; ...
        \end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{The TP53 locus}

        \begin{figure}[!]
            \includegraphics[width=0.95\textwidth]{fig/cdk2na}
        \end{figure}

        \begin{itemize}
            \item Cell cycle regulator
            \item GWAS categories: cancer;cardiovascular disease;hypertensive disorder;joint disease;respiratory system disease;skin cancer
            \item eQTL biosamples: adipose tissue;adrenal gland;aorta;blood;brain; ...
        \end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Web access}

        \begin{figure}[!]
            \includegraphics[width=\textwidth]{fig/web_portal.png}\label{fig:coloc_web}
        \end{figure}

    \section{Concluding remarks} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

        \begin{itemize}
            \item A large set of GWAS variants annotated with putative target genes and tissues
            \item A list of pleiotropic eQTLs and regions
            \item Pleiotropic eQTL region include immune and cell cycle regulators
            \item Enrichment of 5' UTR and transcript (missense) variants
            \item Increased number of associated egenes and etissues
            \item A web site to access the data
        \end{itemize}

@@

        \begin{itemize}
            \item Verify causality of eQTL genes on GWAS
            \item Analyze pleiotropic loci in more detail
        \end{itemize}
